---
output: pdf_document
header-includes:
  - \usepackage{pxfonts} 
params:
  results_path: temp
  country_code: IT 
  results_file: it_66001141_57_mob_results.rds 
  fit_file: it_66001141_57_mob.RData
  subnat_hier: true 
  multi_singletons: false 
  extract_results: false 
  suppress_imputed_cases: true
---

```{r setup, echo=FALSE, include=FALSE}
root_path <- "."

if (!interactive()) {
  source(file.path("renv", "activate.R"))
}

library(magrittr)
library(tidyverse)
library(cowplot)
library(ggridges)
library(haven)
library(latex2exp)
library(rstan)

ggplot2::theme_set(theme_minimal())

knitr::opts_chunk$set(echo = FALSE)
```

```{r}
source(file.path("util.R"))
source(file.path("mobility-model", "mob_util.R"))
```

```{r}
do_diag <- !is_null(params$fit_file)
do_prior_diag <- !is_null(params$prior_fit_file)

results <- read_rds(file.path(params$results_path, params$results_file))

if (!is_null(params$country_code)) {
  results %<>%
    mutate(run_data = map(run_data, filter, fct_match(country_code, params$country_code))) 
}

if (do_diag) {
  load(file.path(params$results_path, params$fit_file))
  
  if (params$extract_results) {
    all_parameters <- extract_parameters(mob_fit)
  
    max_rhat <- all_parameters %>%
      select(rhat) %>%
      summarize_all(max, na.rm = TRUE) %>%
      first()
  
    min_ess_bulk <- all_parameters %>%
      select(ess_bulk) %>%
      summarize_all(min, na.rm = TRUE) %>%
      first()
    
    min_ess_tail <- all_parameters %>%
      select(ess_tail) %>%
      summarize_all(min, na.rm = TRUE) %>%
      first()

    results %<>%
      select(run_data) %>% 
      unnest(run_data) %>% 
      select(-param_results) %>% 
      unnest(country_data) %>% 
      extract_results(mob_fit, 
                      hyper_params = c("toplevel_log_R0"), 
                      nat_params = c("national_log_R0", "national_effect_log_R0", "subnational_effect_log_R0_sd"),
                      subnat_params = c("log_R0", "subnational_effect_log_R0", "imputed_cases", "ifr", "trend_lambda", "trend_kappa"),
                      day_params = c("Rt", "Rt_adj", "adj_factor", "mobility_effect", "mean_deaths", "trend", "new_cases", "deaths_rep")) %>% 
      mutate(
        max_rhat,
        min_ess_bulk,
        min_ess_tail,
        div_trans = get_num_divergent(mob_fit),
      )
  }
}

country_names <- results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  pull(country_name)
```

---
title: `r if (length(country_names) > 1) 'Multiple Countries' else country_names`
---

```{r, results='asis'}
if (length(country_names) > 1) {
  cat("Countries:\n\n")
  cat(str_c("* ", results$country_name, collapse = "\n"))
}
```

# Data

```{r pop, fig.width=8}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_population(at_level = if (params$multi_singletons) "country_name" else "sub_region")
```

```{r obs-cases, fig.width=15}
results %>%
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_data(par = "new_cases", use_date = TRUE, use_free_y_scale = TRUE, use_bar = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = "New Cases") 
```

```{r obs-deaths, fig.width=15}
results %>%
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_data(par = "new_deaths", use_date = TRUE, use_free_y_scale = TRUE, use_bar = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region")  +
  labs(title = "New Deaths") 
```

```{r mob-plot, fig.height=6, fig.width=12}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_data(par = c("g_residential", "g_transit_stations", "average_mob"), use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = "Mobility Data") +
  NULL
```


# Analysis

Number of divergent transitions = `r str_c(results$div_trans, collapse = ', ')`

Maximum $\hat{R}$ = `r str_c(results$max_rhat, collapse = ', ')`

Minimum Bulk ESS = `r str_c(results$min_ess_bulk, collapse = ', ')`

Minimum Tail ESS = `r str_c(results$min_ess_tail, collapse = ', ')`

```{r R0-plot, fig.width=10}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "R0") + 
  labs(title = TeX("R_{0,m}"))
```

```{r Rt-plot, fig.height=6, fig.width=12}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = c("Rt", "Rt_adj"), use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(title = TeX("R_{t,m}")) 
```

Contact rate function: \[ cr(t; t^*, \lambda_j, \kappa) = \lambda_j + \frac{1 - \lambda_j}{1 + \exp(\kappa (t - t^*))} \]
where 
\begin{align*}
\lambda_j &\sim \mathtt{Beta}(3, 1) \\
\kappa &\sim \mathtt{NegHalfNormal}(0, 1).
\end{align*}

```{r param-trend, fig.height=6, fig.width=12}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "trend", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = TeX("Contact Rate")) 
```

```{r mobility-effect-plot, fig.height=6, fig.width=12}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "mobility_effect", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(title = "Mobility effect") 
```

```{r adj-factor-plot, fig.height=6, fig.width=12}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "adj_factor", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = TeX("S_{t,m}")) 
```

Mobility linear model: $\beta_1 \cdot X_\text{residential} + \beta_2 \cdot X_\text{transit} + \beta_3 \cdot X_\text{average}$.

```{r beta, fig.width=10, fig.height=8}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(beta = TRUE) + 
  labs(title = TeX("$\\beta$"))
```

```{r post-deaths, fig.width=10, fig.height=8}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_post_deaths(facet_by = if (params$multi_singletons) "country_name" else "sub_region")
```

```{r imputed-cases, fig.width=10, eval=!params$suppress_imputed_cases}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "imputed_cases") + 
  labs(title = TeX("Imputed Cases")),
```

```{r cases, fig.width=10, fig.height=8}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_pred_vs_obs("new_cases", "new_cases", facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = "New Cases", subtitle = "predicted vs observed")
```

```{r ifr, fig.width=10}
results %>% 
  select(run_data) %>% 
  unnest(run_data) %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "ifr") + 
  labs(title = TeX("IFR"))
```

# Prior Prediction

```{r}
if (!is_null(params$prior_results_file)) {
  do_prior_predict <- TRUE
} else {
  do_prior_predict <- FALSE
  cat("Not Available.")
}
```

```{r prior-predict-setup, include=FALSE, eval=do_prior_predict}
prior_results <- read_rds(file.path(params$results_path, params$prior_results_file))
```

```{r prior-R0-plot, fig.width=10, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "R0", violin_density = TRUE) + 
  labs(title = TeX("R_{0,m}"))
```

```{r prior-Rt-plot, fig.height=6, fig.width=12, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = c("Rt", "Rt_adj"), use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(title = TeX("R_{t,m}")) 
```

```{r prior-param-trend, fig.height=6, fig.width=12, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "trend", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = TeX("Contact Rate")) 
```
```{r prior-mobility-effect-plot, fig.height=6, fig.width=12, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "mobility_effect", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(title = "Mobility effect") 
```
```{r prior-adj-factor-plot, fig.height=6, fig.width=12, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_day_ci(par = "adj_factor", use_date = TRUE, facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = TeX("S_{t,m}")) 
```

```{r prior-beta, fig.width=10, fig.height=8, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(beta = TRUE, violin_density = TRUE) + 
  labs(title = TeX("$\\beta$"))
```

```{r prior-deaths, fig.width=10, fig.height=8, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_post_deaths(facet_by = if (params$multi_singletons) "country_name" else "sub_region")
```

```{r prior-imputed-cases, fig.width=10, eval=do_prior_predict && !params$suppress_imputed_cases}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "imputed_cases", violin_density = TRUE) + 
  labs(title = TeX("Imputed Cases")) +
  coord_cartesian(xlim = c(0, 100))
```

```{r prior-cases, fig.width=10, fig.height=8, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_pred_vs_obs("new_cases", "new_cases", facet_by = if (params$multi_singletons) "country_name" else "sub_region") +
  labs(title = "New Cases", subtitle = "predicted vs observed")
```

```{r prior-ifr, fig.width=10, eval=do_prior_predict}
prior_results %>% 
  select(country_code, country_name, country_data) %>% 
  unnest(country_data) %>% 
  plot_subnat_ci(par = "ifr") + 
  labs(title = TeX("IFR"))
```

# Diagnostics

```{r, results='asis'}
if (!(do_diag || do_prior_diag)) {
  cat("Not Available.")
} else {
  library(bayesplot)
  color_scheme_set("darkgray")
}
```

```{r diag-setup, include=FALSE, eval=do_diag}
mob_posterior <- as.array(mob_fit)

mob_sub_iter <- sample(nrow(mob_posterior), 200) # Shrink it a little

mob_posterior %<>% magrittr::extract(mob_sub_iter, ,)

mob_lp <- log_posterior(mob_fit) %>%
  filter(Iteration %in% mob_sub_iter)
mob_np <- nuts_params(mob_fit) %>%
  filter(Iteration %in% mob_sub_iter)
```

```{r prior-diag-setup, include=FALSE, eval=do_prior_diag}
load(file.path(params$results_path, params$prior_fit_file))

prior_mob_posterior <- as.array(mob_fit)

prior_mob_sub_iter <- sample(nrow(prior_mob_posterior), 200) # Shrink it a little

prior_mob_posterior %<>% magrittr::extract(prior_mob_sub_iter, ,)

prior_mob_lp <- log_posterior(mob_fit) %>%
  filter(Iteration %in% prior_mob_sub_iter)
prior_mob_np <- nuts_params(mob_fit) %>%
  filter(Iteration %in% prior_mob_sub_iter)
```

```{r pairs, fig.width=14, fig.height=10, eval=do_diag}
pairs_pars <- c(
  "overdisp_deaths[1]", "mean_deaths[10]", 
  "toplevel_log_R0", 
  "beta_toplevel[2]", 
  # "ifr_noise[1]",
  "trend_lambda[1]", "toplevel_trend_kappa"
)

if (!params$suppress_imputed_cases) {
  pairs_pars %<>% c("imputed_cases[1]") 
}

pairs_trans <- lst(
  "overdisp_deaths[1]" = "log",
  "mean_deaths[10]" = "log",
  "trend_lambda[1]" = "log", 
  "toplevel_trend_kappa" = function(kappa) log(-kappa),
  # "ifr_noise[1]" = "log"
)

if (!params$suppress_imputed_cases) {
  pairs_trans %<>% update_list(
    "imputed_cases[1]" = "log"
  )
}

if (params$subnat_hier && !params$multi_singletons) {
  pairs_pars %<>% c(
    "subnational_effect_log_R0_raw[1]", "subnational_effect_log_R0_sd[1]",
    "beta_subnational_raw[2,1]", "beta_subnational_sd[2,1]"
  ) 
  
  pairs_trans %<>% list_modify(
    "subnational_effect_log_R0_sd[1]" = "log",
    "beta_subnational_sd[2,1]" = "log"
  )
}

mcmc_pairs(
  mob_posterior, np = mob_np,
  pars = pairs_pars,
  transformations = pairs_trans,
  off_diag_args = list(size = 0.75)
  # np_style = pairs_style_np(div_size = 2, div_shape = 16, td_alpha = 0.1)
)
```

```{r prior-hypers, fig.width=8, eval=do_prior_diag}
# mcmc_areas(prior_mob_posterior, 
#            regex_pars = c("overdisp_deaths"), 
#            prob = 0.8) + ggtitle("Prior predicted distribution of parameters")

mcmc_areas(prior_mob_posterior, 
           regex_pars = c("overdisp_deaths", "beta_toplevel", "beta_national_sd", "toplevel_log_R0", "national_effect_log_R0_sd"), 
           transformations = lst("toplevel_log_R0" = "exp"),
           prob = 0.8) + ggtitle("Prior predicted distribution of parameters")
```

```{r hypers, fig.width=8, eval=do_diag}
# mcmc_areas(mob_posterior, 
#            regex_pars = c("overdisp_deaths"), 
#            prob = 0.8) + ggtitle("Posterior distribution of parameters")

mcmc_areas(mob_posterior, 
           regex_pars = c("overdisp_deaths", "beta_toplevel", "beta_national_sd", "toplevel_log_R0", "national_effect_log_R0_sd"), 
           transformations = lst("toplevel_log_R0" = "exp"),
           prob = 0.8) + ggtitle("Posterior distribution of parameters")
```

