---
title: "Modeling Heterogeneous COVID-19 Spread"
output: 
  bookdown::pdf_document2:
    fig_caption: true
    toc: false
header-includes:
  - \usepackage{pxfonts} 
bibliography: ../references.bib
params:
  results_file: lite_merged_trend.rds
  country_code: BR 
---

```{r setup, echo=FALSE, include=FALSE}
if (!interactive()) {
  old_dir <- setwd("..")
  source(file.path("renv", "activate.R"))
  setwd(old_dir)
}

library(magrittr)
library(tidyverse)
library(cowplot)
library(ggridges)
library(haven)
library(latex2exp)
library(rstan)

ggplot2::theme_set(theme_minimal())

knitr::opts_chunk$set(echo = FALSE)

source(file.path("..", "mobility-model", "mob_util.R"))
```

# Research question

In what follows, we propose a Bayesian model for COVID-19 epidemics. The model incorporates a number of mechanisms that can explain heterogeneous epidemic growth rates and offers a formal, probabilistic way of reasoning about different factors that may drive the heterogeneity. We then attempt to fit the model to national and sub-national data. We use hierarchical model to synthesise evidence across the largest possible number of countries and regions, while making the best possible use of available data through regularising our inference and not discarding the regions which have low spread of COVID.

We can distinguish five main potential mechanisms that would lead to some regions being less affected by the pandemic. In particular, we try to focus on the factors that may be operating in the developing countries. The factors are: 1) low risk of severe outcomes, 2) low seeding, 3) low $R_0$, 4) under-reporting of outcomes, 5) decrease in reproductive number $R$ over time due to behaviour change, government intervention or other factors.

_1) Low risk of severe disease_. While health care quality and utilisation may have an impact on COVID mortality, age is the main determinant of COVID outcomes. In developing countries mortality rates and risk of severe infections will therefore be lower. 

_2) Low seeding_. In deterministic models of infectious diseases, a single case is enough to start the exponential growth in infections. In the real world, however, due to stochastic nature of contacts, more cases may be needed to start an epidemic. We can assume that many places were not (yet) affected by the epidemic due to number of COVID cases at any point has not reached this epidemic threshold. This, in turn, can be due to low connectedness of different regions (especially in developing countries), early government restrictions on travel or simply due to chance.

_3) Low $R_0$_. When reproductive rate of the virus is low, the numbers of cases may either decrease or build up very slowly, with no evident exponential trend for many months. Various factors can explain low $R_0$

* Lower number of super-spreaders or super-spreading events
* Low population density 
* Social norms and social network structure
* Climate (fixed) or weather (time-dependent factors, which will act on $R_t$ - see below)

_4) Under-reporting or other data issues_. In many developing countries deaths due to COVID may either not be attributed to SARS-COV-2 infection or not recorded at all. Even worse, this under-counting cannot be assumed to be constant over time.

_5) Decrease over in $R$ time_. Assuming that the virus was seeded in a particular setting and $R_0$ is sufficiently high (as evidenced by some local outbreaks), the lack of subsequent epidemic could be explained by a few competing hypotheses: 

* Various government interventions designed to slow the spread of the virus through restrictions in activity, detection of cases, tracing of contacts
* Heterogeneous spread of the virus across communities. The higher risk, "higher R" individuals were affected first and the remaining part of the population reproduces the virus at lower rates. Herd immunity can also play a part in this mechanism. This is different from low $R_0$, in that the initial observed infection rate will be high, and only subsequently decrease.
* Change in behaviour in the absence or in addition to government policy, such as can be observed in drops in mobility data in many countries which precede gov't intervention

We hope to distinguish between the five main mechanisms by using the Bayesian hierarchical model. However, many of the possible mechanisms that can lead to no epidemic are not distinguishable from each other in absence of additional information. A simple example is the near reciprocal relationship of $R_0$ and initial seeding. 
For this reason, instead of following a hypothesis-testing approach, we will aim to make a probabilistic argument and make use of informative priors. For example, both $R_0$ and the number of cases early in the epidemic can have strongly informative prior distributions constructed based on travel data, literature on SARS transmission, models of seeding and history of contagion from previous pandemics _etc._ By introducing these additional assumptions we hope to be able to identify the main parameters of the model, thus characterising the main drivers of heterogeneity in the spread of COVID. 



# Data

We currently use data from about 120 countries, 30 of which report COVID outcomes at a sub-national level (usually administrative regions such as states in the USA, lands in Germany, states in India etc.). We arrange our data into a hierarchical structure, but if sub-national data are missing, we only estimate the national-level parameters (see below). When the sub-national estimates are available, the aggregate, national-level data are not used for inference -- national parameters are inferred as Bayesian hyperparameters. 

The inputs for each unit of analysis are: mobility data over time, mortality over time, population size, time to recording the first case, time to first $n$ deaths. (To be expanded.)

For each country we calculate the expected infection fatality rate (IFR) by reweighting the age-specific IFR estimated by @Verity in Chinese individuals through age distribution based on World Population Prospects data.



# Bayesian model of COVID transmission in countries and regions

We extend @Vollmer2020 for our Bayesian modeling framework. The data generating process models the number of infections as latent outcomes, which then predicts the expected number of deaths, which then leads to observed, COVID-attributable deaths. Latent variables are linked through convolutions of random variables. The primary component of the model adds structure to how transmission of infection evolves over time predicted by observed outcomes. Vollmer et al. use Google mobility data as a measure of social contact, which we also do.

First, the transmission of infection $c_{tm}$, in sub-region $m$ on day $t$, is modeled, for $t > 6$, as
\begin{align*} 
  c_{tm} &= S_{tm} R_{tm} \sum_{\tau=1}^{t-1} c_{\tau,m} \tilde{g}_{\tau - 1},  
\end{align*}

and for the first six days we impute the number of cases as

\begin{align}\label{eq:imputed}
  c_{tm} = c_{m}^\textrm{imputed} &\sim \mathtt{Exponential}\left(\frac{1}{\tau^\textrm{imputed}}\right) \\ 
  \tau^{imputed} &\sim \mathtt{Exponential}(0.03)
\end{align}

* $S_{tm} = 1 - \frac{\sum_{i = 1}^{t-1} c_{im}}{N_m}$ is the susceptible proportion of the population, where $N_m$ is the size of the population in sub-region $m$. 
* $\tilde{g}_s$ are a discretization of the a continuous generation \[ g \sim \mathtt{Gamma}(6.5, 0.62), \] such that $g_s = \int_{\min(s-0.5,0)}^{s+0.5} p_g(\tau)\ \mathrm{d}\tau$, where $p_g$ is the density function for $g$.

$R_{tm}$, the dynamic transmission rate is defined as 
\[ R_{tm} = R_{0,m} \cdot \left[ 2 \cdot \phi^{-1} (\mathbf{X}_{tm}\boldsymbol{\beta}_{m})\right] \cdot \rho_{tm}. \]
$R_{0m}$ is modeled as

\begin{align}\label{eq:r0}
  \log R_{0m} &= \log R_0 + \delta_{k[m]} + \delta_m \\
  \log R_0 &\sim \mathtt{Normal}(\mu_{R_0}, \tau_{R_0}) \\
  \delta_{k} &\sim \mathtt{Normal}(0, \tau^{\delta,\textrm{nat}}) \\ 
  \delta_{m} &\sim \mathtt{Normal}(0, \tau^{\delta,\textrm{subnat}}_{k[m]}) \\
  \tau^{\delta,\textrm{nat}} &\sim \mathtt{Normal}^+(0, 0.1) \\
  \tau^{\delta,\textrm{subnat}}_k &\sim \mathtt{Normal}^+(0, 0.8).
\end{align}

We use the subscript $k$ for countries and $m$ for their sub-regions. We use the following hyperparameters: $\tau_{R_0} = \sqrt{\log(\frac{0.25^2}{3.28^2} + 1)}, \mu_{R_0} = \log(3.28) - \frac{\tau_{R_0}^2}{2}$.^[Currently, we are only running countries separately so we are not using the national level effects on $R_0$.] Estimation of this parameter is shown in Figure \ref{fig:R0-plot}.  

$R_t$ is predicted in the inverse logit function by the variables in the design matrix $\mathbf{X}$ such that each parameter in the vector $\boldsymbol{\beta}_m$ is modeled as

\begin{align*}
  \beta_{pm} &= \beta_p + \gamma_{p,k[m]} + \gamma_{pm} \\
  \beta_p &\sim \mathtt{Normal}(0, 0.5) \\
  \gamma_{pk} &\sim \mathtt{Normal}(0, \tau^{\gamma,\textrm{nat}}) \\
  \gamma_{pm} &\sim \mathtt{Normal}(0, \tau^{\gamma,\textrm{subnat}}_{k[m]}) \\
  \tau^{\gamma,\textrm{nat}} &\sim \mathtt{Normal}^+(0, 0.25) \\
  \tau^{\gamma,\textrm{subnat}}_{k[m]} &\sim \mathtt{Normal}^+(0, 0.5).
\end{align*}
Estimation of this parameter is shown in Figure \ref{fig:Rt-plot}.

In addition, we include parametric model of reduced contact rate such that 

\begin{align*}
  \rho_{tm} &= \lambda_m + \frac{1 - \lambda_m}{1 + \exp(\kappa(t - t^*_m))} \\
  \lambda_m &\sim \mathtt{Beta}(3, 1) \\
  \kappa &\sim \mathtt{Normal}^-(0, 1). 
\end{align*}

This is a smoothly decreasing trend in contact representing unobservable qualitative changes in social contact not captured by mobility data. The midpoint of this trend, $t^*_m$, is the day the first covid-19 case is detected. __Work in progress, t* will be defined differently.__

The expected number of deaths, $d_{tm}$ is modeled as

\begin{align}\label{eq:d_tm}
  d_{tm} = \textrm{ifr}^*_m \sum_{\tau = 1}^{t-1} c_{\tau,m} \tilde{\pi}_{t-\tau},
\end{align}

where the infection fatality rate (IFR), $\textrm{ifr}^*_m$, is modeled as

\begin{align*}
  \textrm{ifr}^*_m &= \eta_m \cdot \textrm{ifr}_m \\
  \eta_m &\sim \mathtt{Normal}(1, 0.1).
\end{align*}

In other words we perturb the pre-calculated IFR by a small random variable. Time-to-deaths, $\pi_s$, is modeled as a continuous distribution
\[ \pi_s \sim \mathtt{Gamma}(5.1, 0.86) + \mathtt{Gamma}(17.8, 0.45),\]
which we discretized as $\tilde{\pi}_s = \int_{\min(s-0.5, 0)}^{s+0.5} p_\pi(\tau)\,\mathrm{d}\tau$, where $p_\pi$ is the density function of $\pi$.

Finally, the likelihood model for the observed number of deaths, $D_{tm}$, is 

\[ D_{tm} \sim \mathtt{NegativeBinomial}(d_{tm}, d_{tm} + \frac{d_{tm}^2}{\psi}), \] \label{eq:negbin}

and $\psi \sim \mathtt{Normal}^+(0, 5)$.



## How does the model address the research questions

1. _Low risk of severe disease_: as per (\ref{eq:d_tm}), the model assumes
2. _Low seeding_: this is addressed by (\ref{eq:imputed})
3. _Low $R_0$_: this is addressed by (\ref{eq:r0})
4. _Under-reporting_: this cannot be directly detected by the model, but the likelihood of time-varying under-reporting can be explored through modifying priors and model structure around \ref{eq:negbin}
5. _Decrease in $R$ over time_ is handled by the dynamic component of $R_{tm}$. Here, the impact of various time-dependent explanatory variables can be included, in the same way that mobility currently is. Through $\rho$, we also include a logistic function of time which will lead to monotonic decreases in $R$ that may not be captured by explanatory variables; we expect that $\rho$ is a composite of changes in behaviour or technological changes or decreases in infection force that would be consistent with sub-group heterogeneity. Without extra data it is not possible to distinguish



# Preliminary results -- illustration

```{r results-setup}
results <- read_rds(file.path("..", "data", "mobility", "results", params$results_file)) %>%
  filter(fct_match(country_code, params$country_code))

country_name <- results$country_name %>% unique()
```

In this section we will demonstrate some of the main results of the analysis described here, using data from `r country_name`. While we only show one country, so far this analysis was done on data from 19 countries out of 30.

In Figure \ref{fig:R0-plot}, we summarize the posterior distribution of $R_0$ across all sub-regions. 

```{r R0-plot, fig.width=10, fig.cap="$R_0$ across all sub-regions. Points represent the posterior median, diamonds the posterior mean, and line ranges are the 90% credible intervals."}
gridExtra::grid.arrange(grobs = list(
  results %>% 
    plot_subnat_ci(par = "R0") + 
    labs(title = TeX("R_{0,m}")) +
    theme(strip.text = element_blank()) +
    NULL,
  
  results %>% 
    plot_subnat_ci(par = "imputed_cases") + 
    labs(title = TeX("Number of initially infected (seeding)"))
),
ncol = 2)

```

In Figure \ref{fig:Rt-plot}, we summarize the full time-series of $R_t$'s evolution across all sub-regions.

```{r Rt-plot, fig.height=12, fig.width=10, fig.cap="Adjusted $R_{t}$ across all regions."}
results %>% 
  plot_day_ci(par = c("Rt_adj"), use_date = TRUE) +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(title = TeX("R_{tm}S_{tm}")) +
  theme(axis.text.y = element_text(size = 7)) +
  NULL
```

In Figure \ref{fig:post-deaths}, we compare the model's predicted number of deaths with actual data.

```{r post-deaths, fig.height=12, fig.width=10, fig.cap="Posterior predicted number of new deaths."}
results %>% 
  plot_post_deaths() +
  labs(title = "Posterior predicted number of new deaths.")
```


```{r beta, fig.width=10, fig.cap="Posterior distributions for model coefficients in the mobility model"}
results %>% 
  plot_subnat_ci(beta = TRUE) + 
  labs(title = TeX("$\\beta$"))
```

Mobility linear model is $\beta_1 \cdot X_\text{residential} + \beta_2 \cdot X_\text{transit} + \beta_3 \cdot X_\text{average}$.

\pagebreak

# References {-}
